---
title: "Introduction and Set Up"
editor: source
format: html
toc: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=3.5, fig.align="center")
```


## Learning Objectives

-   Understand the difference between CRAN and Bioconductor
-   Be able to install Bioinformatics packages

## Getting Set Up
RStudio provides a useful feature called Projects which act like a container for your work. As you use R more, you will find it useful to make sure your files and environment for one real-world project are kept together and separate from other projects.

Let’s create a new project now.

1. Go to `File > New Project`
1. In `Create project from` menu choose `New Directory`
1. Choose Project Type `New Project`
1. Make sure `Create project as subdirectory of:` is pointing to Desktop (or whatever your preferred location is)
1. Call your new directory `r_bioinformatics_lesson`
1. Select the check box that says `Open in New Session`
1. Inside your new project, create folders called `data` and `figures`


## What is Bioconductor?

In this lesson, we'll be working with a number of bioinformatics
packages along with the `tidyverse` family of packages. Many R packages
come from **CRAN (Comprehensive R Archive Network)**. Packages from CRAN
can be installed either by using the Install Packages widget in RStudio
(lower-right pane) or with the function `install.packages()`.

[Bioconductor](https://bioconductor.org/) is an open-source project that
provides tools for the analysis and comprehension of high-throughput
biological data, built on the R programming language. It includes a
large ecosystem of packages for tasks such as sequence analysis, genomic
data visualization, and statistical modeling in bioinformatics.
Bioconductor emphasizes reproducibility, interoperability, and the
integration of biological metadata, making it especially well-suited for
omics research.

To use packages from Bioconductor, we must first install and load the
**`Biocmanager`** package from CRAN. `Biocmanager` provides an interface
to the Bioconductor repository.

```{r}
#| eval: false
#| warning: false
#| message: false
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

library(BiocManager)
```

Next, let's install the packages we will be using in this session. From
CRAN, we'll be installing

| Package Name | Purpose                            |
|--------------|------------------------------------|
|`tidyverse`  | Wrangling and visualizing data      |
|`rentrez`    | Accessing data from NCBI databases  |
| `ape`       | Phylogenetic analysis               |


```{r} 
#| eval: false

install.packages(c("tidyverse", "rentrez", "ape"))
```

Then we can install our Bioconductor packages with `Biocmanager`

| Package Name | Purpose                            |
|--------------|------------------------------------|
|`Biostrings`  | Manipulating biological sequences  |
|`pwalign`     | Pairwise Alignment                 |
|`DECIPHER`    | Multiple Sequence Alignment        |

```{r}
#| eval: false

BiocManager::install(c("Biostrings", "pwalign", "DECIPHER"))
```

**NOTE: You only need to install a package once on your system (and after updates)**, but you will want to load the packages into your R session with the `library()` function.

```{r}

#| eval: false

library(ape)
library(Biostrings)
library(DECIPHER)
library(pwalign)
library(rentrez)
library(tidyverse)
```

Don't be afraid if you see a lot of text and messages printing out in the console as you go through this process. You may be asked to update other packages that the ones we are trying to install are dependent on. You will also be warned if some of the packages have functions with the same name.
